Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-08-09 submission/manuscript.Rmd
methods final draft 1 read through

**The number of OTUs positively associated with SCFA concentration were similar between individuals without colon tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without colon tumors and patients with adenomas or carcinomas in SCFA concentrations or genes encoding enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations. The main goal of this analysis was to identify if there were OTUs that were significantly associated with SCFA concentrations and if this was different between individuals that did not have colon tumors, had an adenoma, or had a carcinoma. To accomplish this we used Spearmans rho, a non-parametric measure of association, and tested if there was a correlation that was signficantly greater than zero. We found that taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* dominated statistically significant OTU correlations [Figure 3 & Table S5]. There was a noticeably higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3]. In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas [Figure 3]. Patients with adenomas also had more positive correlations between OTUs and SCFA concentrations, but their total number was more similar to individuals without colon tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3]. Additionally, the number of positive correlations between OTUs and SCFA concentrations was similar between individuals without colon tumors and patients with a carcinoma [Figure 3]. Finally, when we used high/low SCFA groups based on the overall median concentration for each SCFA instead of SCFA concentrations a similar pattern was still observed [Figure S1 & Table S6]. Overall, these results suggest that the resident taxa that may change the most due to colon tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =`r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())` and carcinoma n = `r length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.

@@

**16s rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017]. In brief, sequences were quality filtered and contigs created from the paired end reads. Any sequences with ambiguous base calls were discarded. Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012]. Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptClust algorithm [@opticlust_westcott2017].

@@

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, a Kruskal-Walis rank sum test with a Dunn's post-hoc test was used to assess differences between the groups used. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, the imputed gene counts of important mediators of SCFA synthesis was tested. Additionally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without tumors and patients with adenomas or carcinomas using Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumors using Random Forest models. Finally, models to classify high or low SCFA concentration based on the median of each SCFA or the actual concentration using 16S rRNA gene sequencing data was created using the Random Forest algorithm. For all Random Forest models, the assessment of the most important variables was based on the top 10 features (OTUs or SCFAs) using the mean decrease in accuracy.

>>>>

**The number of OTUs positively associated with SCFA concentration were similar between individuals without colon tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without colon tumors and patients with adenomas or carcinomas in SCFA concentrations or genes encoding enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations. The main goal of this analysis was to identify if there were OTUs that were significantly associated with SCFA concentrations and if this was different between individuals that did not have colon tumors, had an adenoma, or had a carcinoma. To accomplish this we used Spearman's rho, a non-parametric measure of association, and tested if there was a correlation that was signficantly greater than zero. We found that taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* dominated statistically significant OTU correlations [Figure 3 & Table S5]. There was a noticeably higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3]. In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas [Figure 3]. Patients with adenomas also had more positive correlations between OTUs and SCFA concentrations, but their total number was more similar to individuals without colon tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3]. Additionally, the number of positive correlations between OTUs and SCFA concentrations was similar between individuals without colon tumors and patients with a carcinoma [Figure 3]. Finally, when we used high/low SCFA groups based on the overall median concentration for each SCFA instead of SCFA concentrations a similar pattern was still observed [Figure S1 & Table S6]. Overall, these results suggest that the resident taxa that may change the most due to colon tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =`r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())` and carcinoma n = `r length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor (surgery or colonoscopy) with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.

@@

**16s rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017]. In brief, sequences were quality filtered and contigs created from the paired end reads. Any sequences with ambiguous base calls were discarded. Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012]. Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptiClust algorithm [@opticlust_westcott2017].

@@

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, a Kruskal-Wallis rank sum test with a Dunn's post-hoc test was used to assess differences between individuals without colon tumors, patients with adenomas, and patients with carcinomas. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, the imputed gene counts of genes encoding enzymes involved in SCFA synthesis was tested. Additionally, metagenomic sequencing counts for important genes involved with SCFA production were analyzed. From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without colon tumors and patients with adenomas or carcinomas using Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumors using Random Forest models. Finally, models to classify the actual SCFA concentration or high/low SCFA concentration based on the median of each SCFA using 16S rRNA gene sequencing data was created using the Random Forest algorithm. For all Random Forest models, the assessment of the most important variables was based on the top 10 features (OTUs or SCFAs) using the mean decrease in accuracy.

### 2018-08-09 submission/manuscript.tex
methods final draft 1 read through

\textbf{The number of OTUs positively associated with SCFA concentration
were similar between individuals without colon tumors and patients with
adenomas or carcinomas.} Having found no difference between individuals
without colon tumors and patients with adenomas or carcinomas in SCFA
concentrations or genes encoding enzymes involved with SCFA synthesis,
we next investigated if specific OTUs correlated with SCFA
concentrations. The main goal of this analysis was to identify if there
were OTUs that were significantly associated with SCFA concentrations
and if this was different between individuals that did not have colon
tumors, had an adenoma, or had a carcinoma. To accomplish this we used
Spearmans rho, a non-parametric measure of association, and tested if
there was a correlation that was signficantly greater than zero. We
found that taxa from \emph{Clostridiales}, \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} dominated statistically significant OTU
correlations {[}Figure 3 \& Table S5{]}. There was a noticeably higher
number of significant negative correlations associated with patients
with adenomas for all SCFAs tested {[}Figure 3{]}. In particular, OTUs
from the \emph{Ruminococcaceae} family had the largest share of these
negative correlations within patients with adenomas {[}Figure 3{]}.
Patients with adenomas also had more positive correlations between OTUs
and SCFA concentrations, but their total number was more similar to
individuals without colon tumors or patients with carcinomas versus the
analogous comparison for the number of negative correlations {[}Figure
3{]}. Additionally, the number of positive correlations between OTUs and
SCFA concentrations was similar between individuals without colon tumors
and patients with a carcinoma {[}Figure 3{]}. Finally, when we used
high/low SCFA groups based on the overall median concentration for each
SCFA instead of SCFA concentrations a similar pattern was still observed
{[}Figure S1 \& Table S6{]}. Overall, these results suggest that the
resident taxa that may change the most due to colon tumors may not be
ones that are responsible for the production of acetate, butyrate, or
propionate.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (29, 30). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(adenoma n =41 and carcinoma n = 26). For patients undergoing treatment
for their tumor the length of time between their initial and follow up
sample ranged from 188 - 546 days. Our use of treatment has been
previously defined as encompassing removal of a tumor with or without
chemotherapy and radiation (29). Diagnosis of tumor was made by
colonoscopic examination and histopathological review of biopsies
obtained (29, 30). The University of Michigan Institutional Review Board
approved the study and informed consent was obtained from all
participants in accordance to the guidelines set out by the Helsinki
Declaration.

@@

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (29, 31, 32). In brief, sequences were quality
filtered and contigs created from the paired end reads. Any sequences
with ambiguous base calls were discarded. Contigs were then checked for
matches to the V4 region of the 16S rRNA gene using the SILVA database
(33). Chimeras were identified and removed using UCHIME and OTUs
clustered at 97\% similarity (34). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptClust algorithm (35).

@@

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (44). Generally, a Kruskal-Walis rank sum
test with a Dunn's post-hoc test was used to assess differences between
the groups used. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (45). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, the imputed gene counts of
important mediators of SCFA synthesis was tested. Additionally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. From here we analyzed the number of
significant positive and negative correlations between OTU relative
abundance and SCFA concentrations in individuals without tumors and
patients with adenomas or carcinomas using Spearman's rho. Next, we
assessed whether OTUs alone or OTUs and SCFAs were better able to
classify individuals with and without tumors using Random Forest models.
Finally, models to classify high or low SCFA concentration based on the
median of each SCFA or the actual concentration using 16S rRNA gene
sequencing data was created using the Random Forest algorithm. For all
Random Forest models, the assessment of the most important variables was
based on the top 10 features (OTUs or SCFAs) using the mean decrease in
accuracy.

>>>>

\textbf{The number of OTUs positively associated with SCFA concentration
were similar between individuals without colon tumors and patients with
adenomas or carcinomas.} Having found no difference between individuals
without colon tumors and patients with adenomas or carcinomas in SCFA
concentrations or genes encoding enzymes involved with SCFA synthesis,
we next investigated if specific OTUs correlated with SCFA
concentrations. The main goal of this analysis was to identify if there
were OTUs that were significantly associated with SCFA concentrations
and if this was different between individuals that did not have colon
tumors, had an adenoma, or had a carcinoma. To accomplish this we used
Spearman's rho, a non-parametric measure of association, and tested if
there was a correlation that was signficantly greater than zero. We
found that taxa from \emph{Clostridiales}, \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} dominated statistically significant OTU
correlations {[}Figure 3 \& Table S5{]}. There was a noticeably higher
number of significant negative correlations associated with patients
with adenomas for all SCFAs tested {[}Figure 3{]}. In particular, OTUs
from the \emph{Ruminococcaceae} family had the largest share of these
negative correlations within patients with adenomas {[}Figure 3{]}.
Patients with adenomas also had more positive correlations between OTUs
and SCFA concentrations, but their total number was more similar to
individuals without colon tumors or patients with carcinomas versus the
analogous comparison for the number of negative correlations {[}Figure
3{]}. Additionally, the number of positive correlations between OTUs and
SCFA concentrations was similar between individuals without colon tumors
and patients with a carcinoma {[}Figure 3{]}. Finally, when we used
high/low SCFA groups based on the overall median concentration for each
SCFA instead of SCFA concentrations a similar pattern was still observed
{[}Figure S1 \& Table S6{]}. Overall, these results suggest that the
resident taxa that may change the most due to colon tumors may not be
ones that are responsible for the production of acetate, butyrate, or
propionate.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (29, 30). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(adenoma n =41 and carcinoma n = 26). For patients undergoing treatment
for their tumor the length of time between their initial and follow up
sample ranged from 188 - 546 days. Our use of treatment has been
previously defined as encompassing removal of a tumor (surgery or
colonoscopy) with or without chemotherapy and radiation (29). Diagnosis
of tumor was made by colonoscopic examination and histopathological
review of biopsies obtained (29, 30). The University of Michigan
Institutional Review Board approved the study and informed consent was
obtained from all participants in accordance to the guidelines set out
by the Helsinki Declaration.

@@

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (29, 31, 32). In brief, sequences were quality
filtered and contigs created from the paired end reads. Any sequences
with ambiguous base calls were discarded. Contigs were then checked for
matches to the V4 region of the 16S rRNA gene using the SILVA database
(33). Chimeras were identified and removed using UCHIME and OTUs
clustered at 97\% similarity (34). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptiClust algorithm (35).

@@

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (44). Generally, a Kruskal-Wallis rank sum
test with a Dunn's post-hoc test was used to assess differences between
individuals without colon tumors, patients with adenomas, and patients
with carcinomas. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (45). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, the imputed gene counts of
genes encoding enzymes involved in SCFA synthesis was tested.
Additionally, metagenomic sequencing counts for important genes involved
with SCFA production were analyzed. From here we analyzed the number of
significant positive and negative correlations between OTU relative
abundance and SCFA concentrations in individuals without colon tumors
and patients with adenomas or carcinomas using Spearman's rho. Next, we
assessed whether OTUs alone or OTUs and SCFAs were better able to
classify individuals with and without tumors using Random Forest models.
Finally, models to classify the actual SCFA concentration or high/low
SCFA concentration based on the median of each SCFA using 16S rRNA gene
sequencing data was created using the Random Forest algorithm. For all
Random Forest models, the assessment of the most important variables was
based on the top 10 features (OTUs or SCFAs) using the mean decrease in
accuracy.
